---
title: "Getting Started with pakhom"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pakhom}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # All chunks in this getting-started vignette are illustrative -- they show
  # how a user would invoke pakhom against their own data, API keys, and
  # configuration. They are NOT runnable in the package-build environment
  # because they reference user-supplied paths (e.g., "my_data.csv") and
  # require live API credentials. Setting eval = FALSE prevents knitr from
  # running them at vignette-build time; purl = FALSE prevents R CMD check's
  # "running R code from vignettes" step from extracting and running them
  # (Stangle/purl ignores eval = FALSE and respects purl = FALSE instead).
  eval = FALSE,
  purl = FALSE
)
```

## What is pakhom?

**pakhom** is an R package for AI-assisted thematic analysis (after
Braun & Clarke, 2006), offered in three methodologically-distinct modes
(reflexive, codebook, and framework). It takes a dataset of text entries (e.g.,
forum posts, survey responses, interview transcripts) and produces a
publication-quality HTML report containing themes, codes, sentiment analysis,
correlation analysis, and supporting evidence.

The package's name is the Coptic word for *eagle* and honors Saint Pachomius
the Great (c. 292–348 CE), whose written **Rule** of communal discipline
established the genre of methodology-as-written-document. pakhom (the package)
extends that lineage to AI-assisted qualitative analysis: methodology is
codified at the architectural level, not left to user discipline at the
configuration level. See `?pakhom` for the longer note on the name.

You give it your data and your research question. It gives you back a complete
thematic analysis -- including the kinds of tables, figures, and statistical
tests you would normally produce by hand over weeks of manual coding.

### Who is this for?

Researchers who:

- Are conducting qualitative or mixed-methods research with large text datasets
- Want to use AI to assist (not replace) their analytical process
- May or may not have deep experience with R programming
- Want reproducible, transparent, and auditable thematic analysis

### What does it actually do?

The package runs a pipeline that mirrors how manual reflexive thematic
analysis works:

1. **Codebook-first learning** -- Learns from your previous manual analyses
   (QDPX/NVivo codebooks, Excel, CSV) including the full code hierarchy,
   entry-level coding examples, and discarded codes
2. **Data loading** -- Reads from a SQLite database (one or more tables)
3. **Progressive sequential coding** -- AI reads each entry one at a time,
   coding applicable text segments and skipping irrelevant content. Builds a
   growing codebook as it goes (no relevance filter, no batch coding,
   no consolidation step)
4. **Researcher review** (optional) -- You curate the codebook before theming
5. **Code-aware sentiment analysis** -- AI assigns sentiment scores using the
   codes assigned to each entry as context for more accurate scoring
6. **Multi-pass AI theme generation** (Phase 60, the `v2` default) -- the AI
   sees the entire codebook at once and proposes a partition of the codes into
   clusters (pass 1); each further pass takes the prior clusters as units and
   the AI either groups them again or declares the partition converged. There
   are no hardcoded pass counts, cluster-size thresholds, or target theme
   counts -- clustering depth (flat, nested, or anywhere between) is the AI's
   dynamic call for the data at hand. Codes are *grouped*, never combined into
   new codes; theme and subtheme names are assigned in a dedicated labeling
   pass *after* the grouping converges. (The legacy Phase 52 HAC-on-embeddings
   tree-walk is retained as an opt-in `algorithm: "v1"` for back-compatible
   test fixtures.)
7. **Deterministic theme cascading** -- Entry-to-theme assignment flows through
   the code hierarchy (pure R, no AI re-reading of entries)
8. **Correlation analysis** -- Statistical tests between themes and variables
9. **Report generation** -- Produces an interactive HTML report

At multiple points, the pipeline can **pause** and let you review and modify
its decisions before proceeding. You can also **stop and resume** at any
time -- the system tracks which entries have been processed and continues
from where it left off.

---

## Which mode is this vignette for?

pakhom has **three methodology modes** (Reflexive Scaffold, Codebook
Collaborative, Framework Applied). The mode declaration is mandatory in
every config; it shapes what the AI may produce, what the researcher
must author, and which transparency artifacts are mandatory.

**This getting-started vignette covers Mode 2 (Codebook Collaborative)**
— the auto-pipeline you'd recognize from manual codebook TA. The AI
proposes codes + themes; you gate each at pause-points; the package
produces a publication-quality HTML report with full audit trail.

For a worked example of all three modes plus a decision rubric for
choosing among them, see `vignette("methodology-modes")`. In short:

- **Mode 1 (`reflexive_scaffold`)** — AI as Socratic gadfly; researcher
  authors all themes + codes; pakhom contributes provocations only.
  Use [`run_mode1()`](../reference/run_mode1.html). Best for reflexive
  TA + constructionist epistemology.
- **Mode 2 (`codebook_collaborative`)** — AI proposes codes + themes;
  researcher gates each. Use [`run_analysis()`](../reference/run_analysis.html).
  This vignette.
- **Mode 3 (`framework_applied`)** — Apply a researcher-supplied
  theoretical framework (TPB, COM-B, TDF, or your own) verbatim;
  flag anomalies. Use [`run_analysis()`](../reference/run_analysis.html)
  with `methodology.framework_spec_path` set. Best for theory-driven
  / deductive analyses.

---

## Prerequisites

### R and RStudio

You need R (version 4.1 or higher) installed on your computer. We strongly
recommend using [RStudio](https://posit.co/download/rstudio-desktop),
which provides a user-friendly interface for running R code.

If you have never used R before: RStudio has a console pane (usually at the
bottom left) where you type commands and press Enter to run them. That is all
you need to know to use this package.

### An AI API Key

pakhom uses AI services to analyze your text. You need an API key from
one of these providers:

- **OpenAI** (recommended for beginners): Sign up at
  [platform.openai.com](https://platform.openai.com). Go to API Keys and
  create one. It will look like `sk-proj-...`.
- **Anthropic**: Sign up at [console.anthropic.com](https://console.anthropic.com).
  Create an API key. It will look like `sk-ant-...`.

Both services charge per use. A typical analysis of ~500 entries costs
approximately $2--5 USD.

**Important:** Never share your API key or commit it to version control.
The package reads it from an environment variable (explained below).

### Your Data

Your text data must be in a **SQLite database** (a `.db` file). If your data
is currently in CSV, Excel, or another format, you can convert it:

```{r}
# Convert a CSV to SQLite
# install.packages(c("DBI", "RSQLite"))   # if not already installed
library(DBI)
library(RSQLite)

db <- dbConnect(SQLite(), "my_data.db")
my_data <- read.csv("my_data.csv")
dbWriteTable(db, "posts", my_data)
dbDisconnect(db)
```

Each row in your table should represent one text entry (e.g., one forum post,
one survey response). You need at minimum a column containing the text content.

### Optional: Reddit API Access (only if you'll use the built-in scraper)

The package includes an optional Reddit scraper (`scrape_reddit()`) that can
populate your SQLite database directly. **You don't need this if your data
is already prepared.** If you do want to use it, note that as of November 2025
Reddit no longer offers self-service API key creation — you must first be
approved through Reddit's Responsible Builder Program:

  - Review the policy: <https://support.reddithelp.com/hc/en-us/articles/42728983564564-Responsible-Builder-Policy>
  - File a support ticket describing your use case (subreddits, expected
    request volume, whether commercial vs academic). Target turnaround
    is ~7 days for academic / non-commercial research.
  - For sensitive-population subreddits (e.g. anything health-related),
    Reddit reviewers significantly favor applications that cite IRB
    approval or an exemption letter from your institution.
  - Once approved, store credentials in `.Renviron`:
    ```
    REDDIT_CLIENT_ID=your_id_here
    REDDIT_CLIENT_SECRET=your_secret_here
    REDDIT_USER_AGENT=pakhom/1.0.0 (by u/YourRedditUsername)
    ```
  - In your config: `scraping.enabled: true` plus the list of subreddits.

If approval is pending or you'd rather skip the scraper, just leave
`scraping.enabled: false` (the default) and load your own data into a
SQLite `.db` file using whatever method you prefer.

---

## Installation

```{r}
# Install from a local copy of the package:
devtools::install("path/to/pakhom")

# Or install directly from GitHub:
# devtools::install_github("abanoub-armanious/pakhom")
```

If you see errors about missing packages, install them first:

```{r}
install.packages(c("dplyr", "tibble", "stringr", "readr", "jsonlite",
                    "yaml", "DBI", "RSQLite", "httr2", "stringdist",
                    "corrplot", "ggplot2", "rmarkdown", "knitr",
                    "progress", "logger", "tictoc", "rlang"))
```

---

## Step 1: Set Your API Key

The safest way to store your API key is in an `.Renviron` file. This file
lives in your home directory and is read automatically when R starts.

**On Mac/Linux**, open a terminal and run:

```bash
echo 'OPENAI_API_KEY=sk-proj-your-key-here' >> ~/.Renviron
```

**On Windows**, run this in R:

```{r}
# This opens the file for editing
usethis::edit_r_environ()
# Add the line: OPENAI_API_KEY=sk-proj-your-key-here
# Save and restart R
```

Verify it works:

```{r}
Sys.getenv("OPENAI_API_KEY")
# Should print your key (not an empty string)
```

For Anthropic, use `ANTHROPIC_API_KEY` instead.

---

## Step 2: Create Your Configuration File

The configuration file (`config.yaml`) tells pakhom everything it needs
to know about your study. You have three options:

### Option A: Web-Based Wizard (Recommended)

```{r}
library(pakhom)
config_wizard_app("config.yaml")
```

This opens a step-by-step web interface in your browser. It starts with your
**methodology mode** (Reflexive Scaffold, Codebook Collaborative, or Framework
Applied -- the mandatory architectural choice), then walks you through every
other setting with descriptions, validation, and a live YAML preview. When done,
it saves a config that loads cleanly. Requires the `shiny` package
(`install.packages("shiny")`).

### Option B: CLI Wizard

```{r}
config_wizard("config.yaml")
```

A simpler text-based wizard that runs in your R console. Its first prompt is the
methodology mode (and, for Framework Applied, the framework spec); it then asks
for study metadata, data path, and provider.

### Option C: Write it Manually

Create a file called `config.yaml` in your project directory. Here is a
minimal example:

```yaml
study:
  name: "My Research Study"
  research_focus: "How does remote work affect employee wellbeing and work-life balance?"
  research_context: "Online forum discussions"
  concepts:
    - "remote work"
    - "wellbeing"
    - "work-life balance"

ai:
  provider: "openai"
  openai:
    api_key_env: "OPENAI_API_KEY"

data:
  database: "my_data.db"
  tables: "posts"
  source_type: "reddit"

output:
  results_dir: "outputs/results"
  generate_report: true
```

### Understanding the Key Settings

**`study.research_focus`** -- This is the most important setting. It tells the
AI what your research question is. Be specific. Instead of "work stress,"
write "How does remote work affect employee wellbeing and work-life balance?"

**`study.concepts`** -- The key concepts in your study. These help the AI
generate more targeted codes focused on the research question. List 2--5 concepts.

**`ai.provider`** -- Either `"openai"` or `"anthropic"`. OpenAI's GPT-4o is
the default and works well for most studies.

**`data.source_type`** -- Tells the package how to preprocess your data.
Options: `"reddit"`, `"twitter"`, `"clinical"`, `"generic"`. If unsure,
use `"generic"`.

---

## Step 3: Explore Your Data

Before running the analysis, preview what is in your database:

```{r}
library(pakhom)

config <- load_config("config.yaml")
explore_database(config$data$database)
```

This shows you which tables exist, how many rows each has, and what columns
are available. Verify that the text column contains the content you expect.

---

## Step 4: Run the Analysis

```{r}
results <- run_analysis("config.yaml")
```

This runs the full pipeline. Depending on your dataset size, it may take
10--30 minutes. You will see progress messages in the console showing which
step is currently running.

**If something goes wrong mid-run** (network error, computer crash, etc.),
simply re-run with `resume = TRUE`:

```{r}
results <- run_analysis("config.yaml", resume = TRUE)
```

The package saves checkpoints after each expensive step. Resuming skips
already-completed steps and continues from where it left off.

---

## Step 5: Review Your Results

### The HTML Report

The pipeline generates an interactive HTML report at
`outputs/results/<run_id>/analysis_report.html`. Open it in your web browser.
It contains:

- Study overview and methodology description
- Sentiment distribution with histograms
- Full theme descriptions with supporting quotes and code lists
- Correlation matrix and significant findings
- Statistical test results (Mann-Whitney U, Chi-square)
- Detailed per-theme entry tables

### Exported Data Files

All results are also exported as machine-readable files:

| File | Contents |
|------|----------|
| `sentiment_scores.csv` | Every entry with sentiment score, emotion, intensity |
| `codes.csv` | Codebook with all codes and frequencies (Phase 60.6 rename from `consolidated_codes.csv`) |
| `correlations.csv` | All pairwise correlation results |
| `themes.json` | Complete theme definitions (name, description, codes, quotes) |
| `theme_entries/` | One CSV per theme containing its assigned entries |

### Working with Results in R

```{r}
# Access the theme set
ts <- results$theme_set
theme_names(ts)          # List theme names
n_themes(ts)             # Count themes
theme_set_to_tibble(ts)  # Convert to a data frame

# Access the analyzed data (the entries that received codes,
# enriched with sentiment, emotion, and theme columns)
analyzed <- results$analytic_data
# Each row now has columns including:
#   sentiment_score, all_emotions, emotion_intensity,
#   emerged_themes (semicolon-separated theme names for this entry),
#   n_themes (count of themes the entry belongs to),
#   theme_membership_<ThemeName> (one binary 0/1 column per theme,
#                                  for multi-label correlation analysis)

# Access correlations
corr <- results$correlations
significant <- corr[corr$significant == TRUE, ]
```

---

## Researcher Review Points (Human-in-the-Loop)

By default, the pipeline runs end-to-end without stopping. But for rigorous
research, you often want to review and curate the AI's decisions at critical
points. Enable review points in your config:

```yaml
analysis:
  review_points:
    after_coding: true          # Review codebook after progressive coding
    after_themes: true          # Review themes before assignment
```

### How Review Points Work

When a review point is enabled, the pipeline:

1. Exports a CSV to `<output_dir>/researcher_review/`
2. Logs a message telling you which file to review
3. Continues running (so downstream steps still proceed with unreviewed data)

To apply your review:

1. Open the exported CSV in Excel, Google Sheets, or any spreadsheet tool
2. Fill in the action columns (instructions are in the console output)
3. Save the reviewed file with the name indicated (e.g., `codebook_reviewed.csv`)
4. Re-run the pipeline with `resume = TRUE`

The pipeline detects the reviewed file and applies your modifications.

### Review Point A: Codebook (after progressive coding)

After progressive coding completes, the pipeline exports `codebook_review.csv`
with columns: `code_key`, `code_name`, `description`, `frequency`, `n_entries`,
`code_type`, `action`, `new_name`, `merge_into`.

For each code, set `action` to:

- `keep` -- Keep as-is (optionally set `new_name` to rename it)
- `delete` -- Remove this code entirely
- `merge` -- Merge into another code (set `merge_into` to the target code name)
- Leave empty to keep unchanged

### Review Point B: Themes (after theme generation)

After theme generation, the pipeline exports `themes_review.csv` with
columns: `theme_name`, `description`, `codes_included`, `action`, `new_name`,
`new_description`, `merge_into`.

Actions are the same as for codes: `keep`, `delete`, or `merge`.

---

## Learning from Previous Studies

If you have previously completed manual thematic analyses, pakhom can
learn from them to produce more consistent and calibrated results.

### Setting Up

Place your prior analysis manuscripts (DOCX or PDF) in a folder. Each study
should have its own subfolder ending with the word "study":

```
manual analyses/
  dayvigo_study/
    manuscript.docx     <- Your completed analysis
    raw data/           <- Optional: original coded entries
      2024-01-15_user1_Rating 4.5.docx
      2024-02-20_user2_Rating 3.0.docx
  ozempic_study/
    manuscript.docx
    raw data/
```

### Configuring

```yaml
learning:
  enabled: true
  base_dir: "manual analyses"
```

The AI reads the themes and results sections from your manuscripts and uses
them as calibration data. This helps it understand what level of specificity
and depth you expect in themes, and what kinds of codes are relevant to
your research area.

---

## Advanced Configuration

### Full Configuration Reference

```yaml
study:
  name: "Study Name"
  research_focus: "Your research question"
  research_context: "Context description"
  concepts: ["concept1", "concept2", "concept3"]
  researcher_positionality: "Optional: your perspective as researcher"

ai:
  provider: "openai"              # "openai" or "anthropic"
  openai:
    api_key_env: "OPENAI_API_KEY" # Environment variable name
    models:
      primary: "gpt-4o"           # Main analysis model
      fast: "gpt-4o-mini"         # Used for batch operations
    rate_limits:
      rpm: 500                    # Requests per minute
      tpm: 150000                 # Tokens per minute

data:
  database: "data.db"
  tables: "posts"                 # Single table or list of tables
  source_type: "reddit"           # reddit, twitter, clinical, generic
  preprocessing:
    min_char: 50                  # Minimum characters to keep an entry
    # Phase 56: thematic saturation is now AI-arbited; no user knobs.
    # The AI judges saturation per max(20, ceiling(n_corpus / 50))
    # coded entries -- see ?pakhom::run_progressive_coding.

learning:
  enabled: false
  base_dir: "manual analyses"

analysis:
  test_mode:
    enabled: false                # Set true for quick validation runs
    sample_size: 100
    seed: 42

  coding:
    progressive: true             # One entry at a time (sequential coding)
    include_in_vivo: true
    max_retries_per_entry: 1
    checkpoint_interval: 50       # Save progress every N entries

  sentiment:
    code_aware: true              # Use codes as context for sentiment
    batch_size: 20
    dynamic_batching: true

  themes:
    # Per C1 ("AI decides when to stop"), pakhom does not gate theme
    # generation on hardcoded n_themes / pass counts / size thresholds.
    # The default v2 algorithm groups codes across multiple AI passes and
    # stops when the AI declares the partition converged (depth is the
    # AI's dynamic call).
    algorithm: "v2"               # "v2" (default: multi-pass AI clustering)
                                  #   or "v1" (legacy Phase 52 HAC tree walk)
    include_subthemes: true       # Render the emergent subtheme decomposition
    include_quotes: true          # Render representative quotes per theme
    quotes_per_theme: 3
    approach: "inductive"         # "inductive" (Mode 2) or "deductive" (Mode 3)
    max_subtheme_depth: 3         # v1 HAC walker only; no-op under algorithm: "v2"
    max_codes_per_subtheme: 25    # v1 HAC walker only; no-op under algorithm: "v2"
    max_inline_themes: 30         # Phase 58 Tier 5 C-3: top-N inlining cap
    max_inline_themes_temporal: 30  # Phase 58 Tier 5 AH-8: temporal plot

  human_verification:
    enabled: false                # Enable for inter-rater reliability
    sample_size: 20
    seed: 42

  review_points:
    after_coding: false           # Pause for codebook review
    after_themes: false           # Pause for theme review
    format: "csv"                 # "csv" or "qdpx"
    max_iterations: 3             # Cap on recursive review cycles

  correlations:
    method: "spearman"
    adjust_method: "bonferroni"
    min_observations: 30
    min_theme_entries: 5          # Phase 58 Tier 6 H-16: harmonized
    use_multi_label: true
    max_inline_vars: 30           # Phase 58 Tier 5 C-10: lollipop above
    max_inline_themes_network: 30  # Phase 58 Tier 5 AH-9: network top-N

output:
  results_dir: "outputs/results"
  generate_report: true
  generate_correlation_plot: true
  comparison_enabled: true

logging:
  log_level: "INFO"               # DEBUG, INFO, WARN, or ERROR
```

### Using Anthropic Instead of OpenAI

```yaml
ai:
  provider: "anthropic"
  anthropic:
    api_key_env: "ANTHROPIC_API_KEY"
    models:
      primary: "claude-sonnet-4-20250514"
      fast: "claude-sonnet-4-20250514"
```

### Multi-Table Analysis

If your data spans multiple tables (e.g., posts and comments):

```yaml
data:
  database: "data.db"
  tables:
    - "posts"
    - "comments"
```

The package loads and combines them, adding a `source_table` column.

### Overriding Config at Runtime

You can override any config setting without editing the YAML file:

```{r}
results <- run_analysis("config.yaml", config_overrides = list(
  "ai.provider" = "anthropic",
  "analysis.test_mode.enabled" = TRUE,
  "analysis.test_mode.sample_size" = 50
))
```

---

## Comparing Runs

After running the analysis multiple times (e.g., with different configs,
different AI providers, or updated data), you can compare results:

```{r}
runs <- list_available_runs("outputs/results")
comparison <- compare_runs("outputs/results/latest", "outputs/results")
print(comparison)
```

The comparison shows:

- Theme stability across runs (which themes persist, which are new)
- Sentiment distribution changes
- Code evolution (new codes, dropped codes)
- Entry migration between themes

---

## Troubleshooting

### "API rate limit exceeded"

The package handles rate limits automatically with exponential backoff. If
you consistently hit limits, reduce the sentiment batch size:

```yaml
analysis:
  sentiment:
    batch_size: 10    # Default is 20
```

Progressive coding processes one entry at a time, so batch size only applies
to sentiment analysis.

### Pipeline interrupted mid-run

Re-run with `resume = TRUE`. The checkpoint system saves progress after each
expensive step. You will not be charged again for already-completed work.

```{r}
results <- run_analysis("config.yaml", resume = TRUE)
```

### Report fails to render

The HTML report requires `pandoc`. RStudio bundles pandoc automatically. If
running R from the command line, install pandoc separately from
[pandoc.org](https://pandoc.org/installing.html).

### "Column not found" errors

Run `explore_database()` to verify your column names match what the package
expects. The package auto-detects common column patterns, but unusual naming
may require the `source_type: "generic"` setting.

### Test mode for validation

Before committing to a full run, validate your setup with test mode:

```{r}
results <- run_analysis("config.yaml", config_overrides = list(
  "analysis.test_mode.enabled" = TRUE,
  "analysis.test_mode.sample_size" = 20,
  "output.generate_report" = FALSE
))
```

This runs the full pipeline on just 20 entries in a few minutes.

---

## Citing pakhom

If you use this package in published research, please cite:

> [Armanious, A. J.](https://www.linkedin.com/in/abanoubarmanious/) (2026).
> pakhom: AI-Assisted Thematic Analysis with
> Methodology-as-Architecture [R package]. Version 1.0.0.
