---
title: "Methodology Modes: Choosing Mode 1, 2, or 3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methodology Modes: Choosing Mode 1, 2, or 3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # Illustrative chunks that reference user-supplied paths and live API
  # credentials. Setting eval = FALSE prevents knitr from running them
  # at vignette-build time; purl = FALSE prevents R CMD check's
  # "running R code from vignettes" step from extracting and running
  # them (Stangle/purl ignores eval = FALSE and respects purl = FALSE).
  eval = FALSE,
  purl = FALSE
)
```

## Why three modes?

`pakhom`'s architectural commitment **AC1** says: *AI is scaffold by
architecture, not by configuration.* Most "AI for thematic analysis"
tools collapse the methodological question ("what is the AI's role
here?") into a configuration option ("which prompt template do I
want?"). pakhom inverts that: the AI's role is determined by which of
three **methodology modes** you declare, and the package code enforces
the commitments that flow from that mode: what the AI may produce,
what the researcher must author, which transparency artifacts are
mandatory, and which pause-points are required.

The mode declaration is mandatory in every config (no default); it is
locked at run start and stamped onto every output. Any change creates a
**fork run** with `parent_run_id` linkage, so the methodology trail is always reconstructable.

This vignette walks through each mode with a worked example, then
gives a decision rubric for choosing among them.

## Mode 1: Reflexive Scaffold (AI as provocateur)

> *"AI Should Challenge, Not Obey"* (Sarkar 2024, CACM Oct 2024).

In Mode 1, the AI never names themes, codes, or interpretations. The
researcher authors the analytic frame (typically in NVivo, ATLAS.ti,
or MAXQDA). pakhom contributes the **provocateur loop**: five
extractive questioning categories that surface counter-evidence the
researcher's framing might overlook.

### The five provocation categories

| Category | What the AI returns |
|---|---|
| `counter_narrative` | Up to N entries, drawn from the supporting entries plus a bounded sample of non-theme corpus entries shown in the prompt, that frame the construct as not-Y (challenges the theme's framing) |
| `disconfirming_evidence` | Entries from the same prompt context that directly contradict the theme |
| `alternative_interpretation` | Methodologically-defensible alternative theme names that the same supporting quotes could support (without saying which is better) |
| `absent_voice` | Demographic / temporal / linguistic / topical segments of the corpus that are underrepresented in the theme's supporting entries |
| `assumption_surfacing` | Other terms participants use for the same construct + a term the researcher's framing erases |

Each provocation that cites a verbatim quote runs through the same
verification ladder used in Modes 2/3. A
fabricated quote is dropped silently and logged to
`fabrication_log.csv`: fabricated provocations never reach the
researcher.

### Worked example

```{r mode1-example}
library(pakhom)

# 1. Author your themes elsewhere (e.g., NVivo) and load them as a ThemeSet.
#    pakhom never writes themes in Mode 1.
my_themes <- create_theme_set(list(
  list(id = 1, name = "Adoption",
       description = "Researcher-authored: async-communication adoption behaviors",
       codes_included = c("async_routine", "daily_batching")),
  list(id = 2, name = "Resistance",
       description = "Researcher-authored: resistance to the new norms",
       codes_included = c("skips_async", "tool_friction"))
))

# 2. Your corpus -- a tibble with std_id + std_text. theme_membership_*
#    columns indicate which entries support each theme. (Author IDs in
#    std_author drive the participant-spread metric.) In practice
#    you'd build this tibble from a database load + your existing NVivo
#    coding; here a small toy corpus illustrates the shape.
my_corpus <- tibble::tibble(
  std_id   = c("e1", "e2", "e3", "e4"),
  std_text = c(
    "I plan to batch my messages every morning from now on.",
    "My manager encouraged me to protect focus time carefully.",
    "I always slip back into back-to-back calls; the calendar is impossible.",
    "Notification overload makes me drop async habits on busy days."
  ),
  std_author = c("alice", "bob", "carol", "dave"),
  theme_membership_Adoption  = c(1L, 1L, 0L, 0L),
  theme_membership_Resistance = c(0L, 0L, 1L, 1L)
)

# 3. Drive the provocateur loop with full transparency and run-state scaffolding.
result <- run_mode1(
  data        = my_corpus,
  theme_set   = my_themes,
  config_path = "config.yaml",   # methodology.mode = "reflexive_scaffold"
  categories  = c("counter_narrative", "disconfirming_evidence",
                   "absent_voice")  # subset; defaults to all five
)

# result$reflection_log carries:
#   - provocations[]: list of Provocation S3 objects (verified citations)
#   - provocation_attempts: data.frame of every theme x category attempt
#                           (so coverage can distinguish "AI returned 0
#                           legitimately" from "category never attempted")
#   - skipped_themes:  themes the orchestrator bypassed (e.g., zero
#                      supporting entries) with explicit reasons
#   - memos:           reflexive notes (initially empty; you write them)

# 4. Reflexive memos
#    are the shared re-engagement lever, available in every mode. Add memos
#    in response to provocations that move you, then persist them.
result$reflection_log <- add_memo(
  result$reflection_log,
  body = "The AI's counter_narrative for 'Adoption' surfaces e3 + e4
from the candidate sample of non-theme entries. I had coded both under
Resistance, but re-reading them against Adoption: they describe failed
adoption attempts rather than refusal, which suggests Adoption and
Resistance may be one dimension, not two themes. Consider renaming or
merging.",
  type = "theoretical",
  linked_themes = "Adoption"
)
persist_memos(result$reflection_log, result$output_dir)
```

### Production recipe: loading a real corpus from config

The toy corpus above is for illustration. In practice you load a
standardized + preprocessed corpus from your SQLite database via
`load_corpus_from_config()`, then attach the
`theme_membership_*` columns from your external coding workflow
(NVivo / ATLAS.ti / MAXQDA export) before calling `run_mode1()`. The
loader runs the same path that `run_analysis()` uses internally, so
the corpus you give Mode 1 is bit-identical to what Modes 2 and 3
would see.

```{r mode1-production}
library(pakhom)

# 1. Parse the YAML config. methodology.mode must be "reflexive_scaffold".
cfg <- load_config("config.yaml")

# 2. Load + standardize + preprocess the corpus (load_and_combine_tables
#    multi-table path, column detection, std_id/std_text/std_author
#    standardization, text cleaning, and optional test_mode sampling all
#    happen here -- no pakhom::: required).
corpus <- load_corpus_from_config(cfg)

# 3. Attach theme_membership_* columns from your external coding tool.
#    Each column is an integer 0/1 indicator: which entries support
#    which researcher-authored theme. (You export these from NVivo /
#    ATLAS.ti / MAXQDA; pakhom never authors theme membership in Mode 1.)
ids_adoption   <- readLines("nvivo_exports/adoption_ids.txt")
ids_resistance <- readLines("nvivo_exports/resistance_ids.txt")
corpus$theme_membership_Adoption  <- as.integer(corpus$std_id %in% ids_adoption)
corpus$theme_membership_Resistance <- as.integer(corpus$std_id %in% ids_resistance)

# 4. Build the ThemeSet from your researcher-authored themes.
my_themes <- create_theme_set(list(
  list(id = 1, name = "Adoption",
       description = "Researcher-authored: async-communication adoption behaviors",
       codes_included = c("async_routine", "daily_batching")),
  list(id = 2, name = "Resistance",
       description = "Researcher-authored: resistance to the new norms",
       codes_included = c("skips_async", "tool_friction"))
))

# 5. Run the provocateur loop. Pass `config = cfg` (the already-parsed
#    object) to avoid re-parsing the YAML, or pass `config_path =` if
#    you'd rather hand off the path.
result <- run_mode1(data = corpus, theme_set = my_themes, config = cfg)
```

### What you get

A finalized run directory at `outputs/<run-id>_M1/` containing:

- `run_metadata.json`: methodology stamp + mode + run id + framework hash (NA for Mode 1) + timestamp
- `rules/methodology_rules.md`: the AC9 system prompt that governed every AI call
- `reflection_log.json`: full reflection log (provocations + attempts + skips + memos)
- `provocations.csv`: flat exportable provocation list (theme + category + cited quote + verification status)
- `provocation_attempts.csv`: the attempt-tracking matrix that drives the whole-corpus coverage check
- `coverage_mode1.json`: `ProvocationCoverage` (no_silent_skip headline + per-category attempt counts)
- `themes.json`: the researcher-authored themes archived for replay
- `memos/<id>.md`: one Markdown file per memo, with YAML frontmatter
- `fabrication_log.csv`: any fabricated provocation citations dropped during the run
- `ai_decisions.jsonl`: full audit trail of every AI call
- `analysis_report.html`: Mode 1 HTML report with transparency dashboards + per-theme provocations + memo timeline

### When to use Mode 1

- You are conducting reflexive thematic analysis (Braun & Clarke 2022) and want AI to challenge your interpretation, not produce it.
- You have a constructionist or critical-realist epistemology where AI-generated themes would be epistemically incoherent with the methodology.
- Your codebook is small (under ~150 codes) so you can author it manually and want depth over scale.
- You take seriously Vikan et al. 2025's finding that LLMs give only limited support for reflexive TA and that high-quality RTA still requires human interpretive engagement, and you want the package to keep you in the data via provocations rather than hand interpretation to the model.

## Mode 2: Codebook Collaborative (the auto-pipeline)

In Mode 2 the AI proposes codes, then themes, and the researcher
**gates each at pause-points**. This is the workflow most users coming
from a manual coding tradition will recognize: the AI does the
mechanical work, the researcher curates.

### Pause points

By default `run_analysis()` will pause after progressive coding (so
you can review the codebook in CSV form) and again after theme
generation (so you can curate themes before correlations are
computed). At each pause, the pipeline exports a CSV; you edit
in-place and rename it (`codebook_review.csv` to
`codebook_reviewed.csv`); then re-run with `resume = TRUE`.

You can also configure no pauses and let the pipeline run
end-to-end, which is useful for batch and replay runs after the codebook is
stable.

### Worked example

```{r mode2-example}
library(pakhom)

# 1. Create a config (the wizard or programmatically). Methodology
#    declaration is mandatory.
create_config(
  methodology = "codebook_collaborative",
  study_name = "Remote-work wellbeing forum study",
  research_focus = "How do users describe the wellbeing effects of remote work?",
  database_path = "corpus.db",
  output_path = "config.yaml"
)

# 2. Run the full pipeline. Returns invisibly with the analytic_data
#    + theme_set + correlations + insights + paths to the output dir.
result <- run_analysis("config.yaml")

# 3. If pause-points are enabled, run_analysis returns a status message
#    and stops at the pause. Edit the exported CSV and re-run with
#    resume = TRUE:
result <- run_analysis("config.yaml", resume = TRUE)

# 4. The Mode 2 report at result$output_dir/analysis_report.html
#    carries the transparency dashboards + saturation curve + per-theme
#    sentiment breakdown + correlation matrix + AI-synthesis executive
#    summary.
```

### What you get

A finalized run directory at `outputs/<run-id>_M2/` with:

- All universal transparency and run-state artifacts (run_metadata, methodology rules, fabrication_log, ai_decisions)
- `sentiment_scores.csv`: per-entry sentiment + emotions + intensity
- `codes.csv`: the codebook (codes preserved as atomic leaves per C2; renamed from `consolidated_codes.csv`)
- `themes.json`: theme set with merge_history + supporting quotes
- `theme_entries/`: one CSV per theme with member entries
- `correlations.csv`: Spearman correlations + Bonferroni-adjusted p-values
- `analysis_report.html`: Mode 2 HTML report
- `correlation_plot.png`, `theme_network.png`: supporting visualizations

### When to use Mode 2

- You are doing codebook TA, template TA, or any approach where AI-generated codes are an acceptable input to your interpretive work.
- You have a medium-to-large corpus (200+ entries) and the AI's mechanical coding speed is genuinely useful.
- You are willing to gate the AI's output at pause-points rather than accept it wholesale.
- You want IRR + saturation diagnostics and the audit trail as your defensibility argument.

### The v2 theme algorithm

Mode 2's theme-generation step uses the **v2 multi-pass clustering with
label-after-clustering** algorithm
(`R/theme_algorithm_v2.R::generate_themes_multipass`). The AI sees the
entire codebook at once, proposes a partition into clusters, and either
continues with another pass (grouping clusters further) or declares
convergence. After convergence, a dedicated labeling pass with the
whole tree visible assigns researcher-facing names + descriptions.

The algorithm honors C-tenets 1 (AI decides when to stop -- no hardcoded
pass count), 2 (codes preserved as atomic leaves), 3 (live tracking per
pass: `outputs/<run>/live/clustering_pass_<N>.json`), and 5 (label after
clustering, never during).

**Empirical evidence from validation** (Mode 2 on a 250-entry
online forum corpus):

| Run | Research focus | Codes | Substantive passes | **v2 themes** | Single-code | **v1 baseline** |
|-----|---|---:|---:|---:|---:|---:|
| 1 | narrow three-concept intersection | 40 | 1 | **6** | 0% | 69 (87% single-code) |
| 2 | broader single-concept focus | 47 | 1 | **10** | 0% | 117 |
| 6 | affect-oriented focus | 157 | **3** | **7** | 0% | 154 (92% single-code) |

In these three validation runs (n = 3 -- indicative, not a performance
guarantee) all landed in a 4-10 theme range with zero single-code themes.
The Run 6 case (157 codes, then 3 substantive passes, then 7 themes with 2-3
subthemes each) demonstrates that v2 can produce a clean theme/subtheme
hierarchy at realistic codebook scale. Single-code reduction is
run-dependent, not a guaranteed invariant (C-tenet 1: the AI judges
convergence; there are no count thresholds).

### Dynamic clustering depth, the AI analyst, and the honesty layer

**Clustering depth is the AI's call, per study.** The number of passes and the
depth of the theme/subtheme hierarchy are emergent, not configured: a narrow,
cohesive codebook may converge in a single pass to a handful of flat themes,
while a broad codebook may take several passes into a deeper hierarchy. Flat
output is not a defect and deep output is not a goal -- both are valid readings
of different data (C-tenet 1). pakhom never imposes a target shape, and it never
combines codes into new codes: every pass returns a *partition* of the existing
codes, and a theme is the *union* of the original codes grouped under it
(C-tenet 2). The themes it produces are a refinable scaffold -- enable the
`after_themes` review point to inspect and adjust the grouping before the report
is built.

**Named facets may disperse (Mode 2) or be guaranteed (Mode 3).** In inductive
Mode 2, a secondary facet you named in your research focus (say, "physical
effects" within a broad lived-experience study) may not surface as its own
theme: the codes that express it can legitimately disperse across the emotional
and behavioral themes where they cohere more strongly. That is sound emergent
grouping -- the content is preserved in the codes and visible in the per-theme
breakdowns -- not a coverage failure. The Mode 2 report includes a
**research-question coverage** section that maps each named facet to where it
landed across the themes (marking dispersion explicitly as the valid outcome it
is), so you can confirm your study addressed each facet without forcing a theme
per facet. If you need *guaranteed* coverage of named facets, use Mode 3
(Framework Applied), where the facets are the framework constructs and every
construct is reported.

**The AI is the analyst; the package is its calculator.** Before
coding, a Methodology Assistant articulates a relevance criterion (injected into
the coding prompt to keep coding on-focus) and, for each numeric or timestamp
column, decides which computational primitives are an *honest* summary -- a
right-skewed count is summarized by a median and tail measures, not a mean+SD; a
bounded ratio is not treated as an unbounded count. These choices are free-form
(the AI is never handed a fixed menu to pick from) and archived, so the
methodology decisions can be re-applied deterministically on a re-run (via
`config$study$inferred_methodology`, which removes those AI calls from the loop)
-- though the coding, sentiment and synthesis steps still query the model and
are not bit-reproducible.

**The honesty layer.** The report is built to be trustworthy under review:

- *Metric provenance.* Each metric is judged as a substantive measure of the
  phenomenon or as incidental source/platform metadata (upvotes, comment counts),
  and grouped accordingly, so platform reception is never silently read as
  prevalence or severity.
- *Small-n reliability.* Spread and distribution-shape statistics on small
  subthemes are marked as indicative: the analyst supplies a per-column
  reliability floor (`min_reliable_n`) and the per-subtheme table flags any
  spread/shape cell below it. Nothing is hidden -- the value and its n are always
  shown; the threshold is the analyst's judgement, not a fixed cutoff.
- *No circular findings.* A correlation between two AI codings of the same text
  (an affect score and a theme-membership indicator) is internal coding
  consistency, not an empirical association; such pairs are excluded from the
  findings (and from the correlation plot's significance markers) but kept in the
  exported matrix with an `exclusion_reason`, so the exclusion is auditable.
- *Saturation, honestly labelled.* The coverage report distinguishes the entries
  coded, examined, and sampled, and the AI saturation arbiter judges convergence
  from the entries-coded growth curve rather than a fixed heuristic.

None of these introduce user-facing hardcoding: the AI judges per run, and the
package renders those judgements transparently rather than classifying your data
for you.

### Choosing the right configuration

Once you've picked Mode 2 via `methodology_decision_aid()`, the
`configuration_selection_aid()` function helps you set expected outcomes
+ recommended review points based on your corpus shape:

```{r mode2-config-aid, eval=FALSE}
library(pakhom)

# Narrow intersection focus, 250-entry corpus:
configuration_selection_aid(
  mode = "codebook_collaborative",
  corpus_size = 250,
  focus_shape = "narrow_intersection"
)
#> $expected_themes:    [5, 8]
#> $expected_passes:    1
#> $recommended_review_points$after_coding: TRUE
#> $expected_wall_time_min: 11
#> $expected_api_spend_usd: 3
```

The bracket is **indicative, not a guarantee** -- it is derived from only
three validation runs (above), so treat the numbers as rough
expectations that will vary with your corpus and model. A narrow research
focus tends to produce fewer codes per coded entry (more cohesive); a broad
focus produces more codes and may require more clustering passes.

## Mode 3: Framework Applied (apply a theoretical framework verbatim)

In Mode 3 you supply a theoretical framework (e.g., the Theory of
Planned Behavior, COM-B, or the Theoretical Domains Framework). The AI
applies it verbatim, coding entries with the framework's constructs as
labels, and flags entries that **resist** the framework as anomalies
per the framework's anomaly_handling policy.

### Built-in frameworks

```{r mode3-frameworks}
library(pakhom)
list_builtin_frameworks()
#> [1] "tpb"  "comb" "tdf"

# Each is loadable by alias OR by file path
spec <- load_framework_spec("tpb")
print(spec)
#> FrameworkSpec: Theory of Planned Behavior
#>   Epistemic stance: positivist
#>   Anomaly policy:   bracket
#>   Constructs:       5
#>     - attitude (Attitude toward the behavior): ...
#>     - subjective_norm (Subjective norm): ...
#>     - perceived_behavioral_control (Perceived behavioral control): ...
#>     - intention (Behavioral intention): ...
#>     - behavior (Behavior): ...
#>   Citations:        2
```

### The three anomaly_handling policies

What happens to entries that don't fit any framework construct depends
on the framework's `anomaly_handling` field:

| Policy | Behavior |
|---|---|
| `bracket` | Entry coded as `anomaly` with a one-sentence reason; framework NOT modified. Most positivist. |
| `extend` | Anomaly becomes a new construct (Vila-Henninger 2024 "abductive coding"); requires explicit researcher acceptance. |
| `revise` | Anomaly triggers modification of an existing construct's definition; logged as framework revision. |

### Worked example

```{r mode3-example}
library(pakhom)

# 1. Create a Mode 3 config -- framework is applied verbatim.
create_config(
  methodology = "framework_applied",
  framework_spec_path = "tpb",   # built-in alias OR path to your YAML/JSON
  study_name = "TPB analysis: async-communication adoption",
  research_focus = "Behavioral intention -> async-communication behavior",
  database_path = "corpus.db",
  output_path = "config.yaml"
)

# 2. Run the pipeline. Internally:
#    - load_framework_spec() -> validates the spec
#    - archive_framework_spec() -> writes outputs/<run>/framework_applied.yaml
#                                  + sha256 hash for replay-equivalence
#    - run_metadata.json carries framework_name + framework_hash +
#      framework_epistemic_stance + framework_anomaly_handling +
#      framework_n_constructs
#    - The Mode 3 HTML report renders a Framework Declaration section
#      with the framework's name + sha256 + citations + epistemic stance +
#      anomaly handling policy + full constructs table
result <- run_analysis("config.yaml")

# 3. The report at result$output_dir/analysis_report.html includes a
#    new "Theoretical Framework (Mode 3 / AC4)" section. The
#    framework_applied.yaml file alongside is byte-equivalent to the
#    spec that was loaded -- a downstream replay using the same hash
#    is provably the same framework.
```

### A note on Mode 3 + Anthropic

When `mode = "framework_applied"` AND `provider = "anthropic"`, the
Anthropic Citations API path is structurally precluded: forced
`tool_use` schema (which Mode 3 requires to constrain coding to
framework constructs) and the Citations API output format are mutually
exclusive on the same response. The Mode 3 + Anthropic pipeline
therefore relies on the verification ladder's DETECTION-only path
(model_freeform + offline string match) rather than the API's
PREVENTION layer.

The Mode 3 report renders an **explicit footnote** disclosing this
constraint. Without the footnote, a reviewer reading "Model freeform
(detection only)" in the transparency dashboard would reasonably wonder why
the Anthropic prevention layer isn't engaged, so the footnote makes the
architectural reason explicit rather than letting it look like a bug.

A hybrid schema (constrained constructs paired with citation
offsets) is a possible future direction.

### When to use Mode 3

- You have a pre-existing theoretical framework you want to apply rigorously.
- You are conducting deductive coding, content analysis, or framework analysis (Ritchie & Spencer).
- The framework's epistemic stance and anomaly handling policy are
  intentional methodological choices, not defaults.

## Choosing among the three modes

| You are doing... | Mode |
|---|---|
| Reflexive thematic analysis (Braun & Clarke 2022); want AI to challenge, not produce | **1** |
| Codebook TA / template TA; want AI mechanical coding with researcher curation | **2** |
| Theoretical framework analysis (TPB, COM-B, TDF, your own) with deductive coding | **3** |
| Constructionist / critical realist epistemology; AI-as-author would be incoherent | **1** |
| Positivist or pragmatic epistemology; AI-as-coder is acceptable input | **2** or **3** |
| Small corpus (under ~150 codes); want depth | **1** |
| Medium-to-large corpus (200+ entries); want scale + audit trail | **2** |
| Pre-registered framework analysis | **3** |
| Your research focus differs from the corpus's *dominant signal* | **1** (see "Mode 2 drift" below) |

## Mode 2 drift on skewed-signal corpora

A re-validation on a 9,178-entry online discussion-forum corpus
surfaced a recurring failure mode you should know about before
choosing Mode 2:

**When the corpus's dominant signal is NOT the configured research focus,
Mode 2 themes drift toward the corpus's natural topic structure and away
from the question you asked.**

Concretely: the run configured a narrow three-concept research
focus against a large general-interest forum's posts and comments. The
corpus's actual dominant signal was the forum's broad everyday affect, not
that narrow intersection. Mode 2's
earlier clustering algorithm faithfully recovered the dominant signal --
417 themes, organized cleanly. But of those 417, only **2** named all
three target concepts together, and only **1** had
substantive mass (~56 entries). The load-bearing three-way
interaction finding was buried in the long tail.

(Note: the algorithm has since been rewritten as multi-pass
AI clustering with label-after-clustering, which surfaces a handful of
themes rather than hundreds of single-code buckets. The narrative above
describes a run of the earlier algorithm; under the current algorithm, the
niche-intersection findings should surface as their own theme rather than
vanishing in 417 single-code buckets. The indicative re-validation (n = 3,
summarized earlier in this vignette) found the current algorithm does this;
broader validation remains future work.)

This is not a bug in Mode 2 fundamentally. Any code-similarity-based
clustering -- HAC, multi-pass AI partitioning, or other -- is
bottom-up: it clusters by code-code similarity, not by research-
question relevance. If 80% of the corpus is about one dominant topic,
80% of the themes will likely be about that topic, regardless of
what your `research_focus` says.

### What to do about it

1. **Audit your corpus's signal balance first**. If `research_focus`
   is a niche intersection (e.g., a specific three-concept intersection),
   sample 100-200 random entries and ask: how many actually touch ALL
   three concepts? If the answer is "fewer than 10%", Mode 2 will
   produce themes that *contain* your focus but won't be *organized
   around* it.

2. **Switch to Mode 1 (Reflexive Scaffold) for the niche-focus case**.
   In Mode 1 you author the themes; the AI doesn't get a vote on
   theme generation. The themes are guaranteed to be organized around
   your `research_focus`, and the AI's role is restricted to
   provocateur loops + memo prompts that challenge what you wrote.
   See `vignette("getting-started")` for the Mode 1 walkthrough.

3. **Refine `research_focus` to match the dominant signal**.
   If your data is mostly about one dominant topic, you can rewrite
   the question as "How does that dominant topic intersect with my
   secondary concepts when it does?" The themes will then be about
   the dominant topic with your secondary concepts as discriminating
   features -- a publishable framing.

4. **Hybrid pipeline (intermediate effort)**. Run Mode 2 first, then
   manually curate the long-tail themes that touch your niche focus
   into a Mode 1 frame for a focused write-up. The Mode 2 audit log
   gives you cross-theme traceability for that curation.

Mode 2 is excellent when the research focus matches the corpus's
dominant signal (e.g., a topic-focused forum analyzed with a research
focus on that same topic).
It is not the right tool for niche-intersection focuses on broad
corpora.

## What every mode produces (the universal transparency layer)

Regardless of mode, every finalized pakhom run produces these
**universal transparency artifacts** (per AC7):

- **Quote provenance (T0.1).** Every AI-attributed verbatim claim is verified through the verification ladder against the cleaned analytic text the model coded (the raw platform text is preserved in `original_text`), and fabrications are dropped and logged to `fabrication_log.csv`.
- **Participant spread (T0.2).** Every theme reports n_distinct_contributors, a Gini coefficient, and the top contributor's share. A theme that looks prevalent but rests on one heavy poster gets a warning on the report.
- **Whole-corpus coverage (T0.3).** Modes 2/3 assert that every preprocessed entry reached the LLM (entry-level coverage; entries over the per-entry character cap are sent truncated, with the truncation measured and disclosed on the coverage card). Mode 1 asserts that there is no silent skip across themes and provocation categories. Both are surfaced via a coverage card on the HTML report.

These commitments are load-bearing: a finalized pakhom run that lacks
any of them is a transparency failure surfaced by
`verify_run_integrity()`.

## Replay-equivalence and run_metadata.json

Every finalized run carries `run_metadata.json` with:

- `run_id` + `methodology_mode` + `mode_locked_at` + `is_finalized`
- `provider` + `model_primary` + `model_fast` (which AI ran the analysis)
- `config_hash` (so a config drift between runs is detectable)
- `framework_name` + `framework_hash` + `framework_*` (Mode 3 only)
- `mode1_categories_requested` + `mode1_n_themes_input` (Mode 1 only)
- `parent_run_id` + `mode_changed_from` (when the run was forked from another)
- `analysis_schema_version` (output column schema; `compare_runs()` uses this to refuse cross-schema comparisons)

This metadata is the contract that makes pakhom runs **auditable and
comparable across re-runs**: two runs against the same data + config +
framework hash + provider produce comparable (not bit-identical) artifacts. Cross-mode comparisons via `compare_runs()`
or `compare_models()` route off these fields.

## The two commitment families (AC1-AC10 and C1-C8)

The package commits to two distinct families of architectural promises:

- **AC1-AC10 (mode-design commitments)** govern *what modes are* and
  *how methodology is enforced*. They are the answer to "why three
  modes? why is there no default?" Documented in the
  README's "architectural commitments" section.
- **C1-C8 (rewrite-direction commitments)** govern *how the coding,
  clustering, and statistics algorithms think*. They are the answer to
  "why doesn't the AI just stop wherever the prompt says?" Surfaced
  in the README's "rewrite-direction commitments" section.

The eighteen commitments together are the load-bearing contract.
Every methodologist reading the package's claims should be able to
verify both sets against the code; anyone
touching the algorithm layer should re-read both before changing
behavior. The two families are orthogonal: mode-design changes
(adding a fourth mode, say, which AC2 forbids) and algorithm
changes (replacing single-pass HAC with multi-pass clustering) are independent.

For algorithm authors specifically, the C-commitments to watch:

| Commitment | Where it lives in code |
|---|---|
| C1: AI decides when to stop | `R/saturation_arbiter.R`, `R/theme_algorithm_v2.R` (multi-pass convergence) |
| C2: Codes preserved through clustering | `R/12_theme_data.R` (Code S3), `R/theme_algorithm_v2.R::apply_partition` (key flattening only, no name mutation) |
| C3: Live tracking artifacts | `R/live_tracking.R` |
| C4: Dataset-agnostic | `R/16_report_helpers.R::.detect_metric_columns`, `R/14_correlations.R` |
| C5: No catch-all buckets | `R/methodology_rules.R` (inductive coding rules) |
| C6: Arbitrary research-question complexity | `R/01_config.R` (no length validation on research_focus) |
| C7: Mode-aware behavior | `R/18_pipeline.R`, `R/mode1_orchestrator.R`, framework-spec handling |
| C8: Publication-quality output shape | `R/16_report_helpers.R::.build_subtheme_summary_table`, `R/17_report.R` |

If you contribute and aren't sure whether a change violates a
commitment, the rule of thumb is: **does this change introduce a
hardcoded number, a hardcoded column name, or a hardcoded mode
assumption?** If yes, you're probably violating C1, C4, or C7. Open a
PR with the commitment cited and discuss before merging.

## Further reading

- `vignette("getting-started")`: full step-by-step Mode 2 walkthrough
- `?run_mode1`: Mode 1 orchestrator API reference
- `?run_analysis`: Mode 2/3 orchestrator API reference
- `?load_framework_spec`: framework spec loading + built-in frameworks
- `?add_memo`: Mode 1 reflexive memo CRUD
- `?compute_mode1_coverage`: Mode 1 T0.3 coverage compute
- Sarkar 2024 (CACM, Oct 2024), "AI Should Challenge, Not Obey": Mode 1 motivation
- Jowsey et al. 2025 (PLOS One, doi:10.1371/journal.pone.0330217), the "Frankenstein" finding: motivation for the transparency layer
- Braun & Clarke 2022: reflexive TA foundation
- Vila-Henninger 2024: abductive coding (Mode 3 `extend` policy)
- Lin & Corley 2025 (arXiv:2505.03105): methodology rules pattern (AC9)
